<!-- REVISED RESULTS — prediction-validation structure for PNAS -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- SI Appendix figure numbering (supplement.Rmd order):
     S1  fig-converge       convergence trajectories
     S2  fig-sim-null-mixed null and mixed simulation scenarios
     S3  fig-cindex-by-k    C-index vs K sensitivity
     S4  fig-k3-k7-nesting  K=3 vs K=7 factor correspondence
     S5  fig-cutpoint-km    cutpoint selection and KM curves
     S6  fig-subtype-overlap subtype composition of risk groups
     S7  fig-varsurvival    variance explained vs survival contribution
     S8  fig-corr           NMF vs DeSurv factor W-matrix correspondence
     S9  fig-forest         per-cohort hazard ratio forest plot
-->

```{r setup}
library(targets)
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(enrichplot)
library(magick)
library(Seurat)
library(gt)
library(VAM)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)


tar_load(sim_figs_by_scenario)

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

alt_plots = lapply(alt_plots, set_fig_font, size = 10)
null_plots = lapply(null_plots, set_fig_font, size = 10)

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
mixed_plots = lapply(mixed_plots, set_fig_font, size = 10)
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Variance vs survival statistics for PDAC training data ───────────────────
tar_load(fig_variation_explained_tcgacptac)
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)

# ── Pooled PDAC validation statistics ────────────────────────────────────────
tar_load(val_latent_desurv_tcgacptac)
pool_dfs <- lapply(val_latent_desurv_tcgacptac, function(entry) {
  surv <- entry$survival
  if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
  data.frame(time = surv$time, event = surv$event,
             dataset = entry$dataset, risk_score = entry$risk_score)
})
pool_df      <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
pool_df      <- pool_df[is.finite(pool_df$time) & is.finite(pool_df$risk_score), ]
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))
cox_pool     <- coxph(Surv(time, event) ~ risk_score + strata(dataset), data = pool_df)
cox_pool_sum <- summary(cox_pool)
hr_pool      <- round(cox_pool_sum$coefficients[, "exp(coef)"], 2)
ci_pool_lo   <- round(cox_pool_sum$conf.int[, "lower .95"], 2)
ci_pool_hi   <- round(cox_pool_sum$conf.int[, "upper .95"], 2)
p_pool       <- signif(cox_pool_sum$coefficients[, "Pr(>|z|)"], 2)

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- tar_read(tar_k_selection_tcgacptac)
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- tar_read(desurv_bo_results_tcgacptac)$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)
```

## Model Overview {.unnumbered}

We developed DeSurv, a survival-supervised deconvolution framework that jointly optimizes NMF reconstruction and Cox proportional hazards likelihood (Fig. \ref{fig:schema}; Methods). The survival gradient acts on the gene program matrix $W$ but not on the sample loadings $H$, directing gene programs toward outcome-relevant structure while preserving interpretability of sample loadings as mixture coefficients [@lee1999learning; @gaujoux2010flexible]. We evaluated DeSurv in two settings: controlled simulations and PDAC cohort analysis with external validation.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..","figures", "model_schematic_final.pdf"))
```


## Survival supervision clarifies NMF rank selection {.unnumbered}

Before examining DeSurv's performance across the two evaluation settings described above, we address a prerequisite question using both PDAC expression data and controlled simulations: whether incorporating clinical outcomes clarifies the well-documented instability of unsupervised rank selection heuristics [@frigyesi2008nmf; @brunet2004metagenes]. Using gene expression data from the PDAC training cohorts (TCGA [@raphael2017integrated] and CPTAC [@ellis2013clinical]; Methods), we first evaluated commonly used unsupervised criteria across a range of candidate ranks.

Standard NMF diagnostics yielded inconsistent guidance (Fig. \ref{fig:bo}A-C). Reconstruction residuals decreased smoothly with increasing $k$ and did not exhibit a clear elbow, a pattern consistent with both relatively small solutions ($k \approx 3$-4) and substantially larger ranks ($k \approx 6$-8). The cophenetic correlation coefficient began to decline at low ranks ($k \approx 3$-4) but continued to fluctuate at higher values without a distinct transition point. In contrast, mean silhouette width was highest at very small ranks ($k \approx 2$-3) and decreased monotonically thereafter, favoring low-dimensional solutions that conflicted with the other criteria. Together, these unsupervised heuristics pointed to incompatible values of $k$, illustrating the ambiguity of rank selection in standard NMF.

To address this ambiguity, we applied DeSurv, which incorporates survival outcomes directly into the factorization and evaluates models using cross-validated concordance index (C-index). The resulting C-index surface across the joint space of factorization rank ($k$) and supervision strength ($\alpha$) identified a coherent region of high performance, in contrast to the conflicting recommendations produced by unsupervised heuristics (Fig. \ref{fig:bo}D). Model selection followed the one-standard-error rule (Methods): we selected the smallest $k$ whose predicted performance lay within one standard error of the maximum, yielding a parsimonious choice. Applying the 1-SE rule to the Bayesian optimization surface selected $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$ and $\alpha = `r round(tar_read(tar_params_best_tcgacptac)$alpha, 3)`$ (peak C-index $`r round(tar_read(tar_k_selection_tcgacptac)$best_mean, 3)`$; $n = `r ncol(tar_read(tar_data_filtered_tcgacptac)$ex)`$ patients, $`r sum(tar_read(tar_data_filtered_tcgacptac)$sampInfo$event)`$ events).

To further evaluate rank recovery under controlled conditions, we conducted simulation studies in which the true underlying rank was known ($k = 3$). DeSurv selected the true rank ($k = 3$) more frequently than standard NMF and avoided the systematic under-selection seen in the unsupervised baseline, though variability across replicates remained (Fig. \ref{fig:bo}E). In contrast, standard NMF followed by post hoc Cox modeling ($\alpha = 0$) showed a dominant tendency to under-select, concentrating near $k = 2$. These results are consistent with the expectation that incorporating outcome information into the factorization itself improves recovery of the true factorization rank.

```{r fig-bo, fig.width = 6, fig.height= 5.5, fig.cap = "(A--D) Analyses based on pancreatic ductal adenocarcinoma (PDAC) gene expression data from TCGA and CPTAC cohorts. (A--C) Standard unsupervised rank selection heuristics yield inconsistent guidance for selecting the factorization rank $k$. (A) Reconstruction residuals as a function of $k$. (B) Cophenetic correlation coefficient as a function of $k$. (C) Mean silhouette width across multiple distance metrics as a function of $k$. (D) Gaussian process predicted mean cross-validated concordance index (C-index) from Bayesian optimization over the joint space of factorization rank ($k$) and supervision strength ($\\alpha$). (E) Simulation studies with known rank ($k = 3$): distribution of selected $k$ values across repeated replicates for DeSurv versus standard NMF with post hoc Cox modeling ($\\alpha = 0$). \\label{fig:bo}", fig.env='figure*', fig.pos='t', out.height= "5.5in", out.width="6in"}

tar_load(fig_bo_heat_tcgacptac)
tar_load(fig_residuals_tcgacptac)
tar_load(fig_cophenetic_tcgacptac)
tar_load(fig_silhouette_tcgacptac)
tar_load(fit_std_tcgacptac)

fig_bo_heat_tcgacptac = set_fig_font(fig_bo_heat_tcgacptac, size = 10)
fig_residuals_tcgacptac = set_fig_font(fig_residuals_tcgacptac, size = 10)
fig_cophenetic_tcgacptac = set_fig_font(fig_cophenetic_tcgacptac, size = 10)
fig_silhouette_tcgacptac = set_fig_font(fig_silhouette_tcgacptac, size = 10)

# upper = plot_grid(fig_bo_cvk_tcgacptac,fig_bo_cvalpha_tcgacptac,labels=c("A","B"))
legend_plot <- plot(fit_std_tcgacptac) + ggplot2::theme(legend.position = "bottom")
legend_grob = ggplotGrob(legend_plot)

legend = legend_grob$grobs[
    sapply(legend_grob$grobs, function(x) x$name) == "guide-box"
][[1]]

upper_panels = plot_grid(
  fig_residuals_tcgacptac  + ggplot2::theme(legend.position = "none"),
  fig_cophenetic_tcgacptac + ggplot2::theme(legend.position = "none"),
  fig_silhouette_tcgacptac + ggplot2::theme(legend.position = "none"),
  ncol = 3,
  labels = c("A","B","C")
)
upper = plot_grid(upper_panels, legend, nrow = 2, rel_heights = c(1, 0.15))

k_hist = plot_grid(alt_plots$k_hist,NULL,nrow=2,rel_heights = c(4,1))

lower = plot_grid(fig_bo_heat_tcgacptac,
                  k_hist,ncol=2,labels=c("D","E"),rel_widths = c(3,2))

plot_grid(upper,lower,nrow=2,rel_heights = c(2,3))

```


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

Sufficient dimension reduction theory predicts that outcome-guided subspace estimation recovers directions most relevant to the response [@cook2007fisher], and related supervised methods have demonstrated empirical benefits in principal component settings [@bair2006supervised], but this prediction has not been evaluated in the NMF deconvolution setting. To test whether it holds under nonnegative constraints with survival outcomes, we designed simulations with known ground-truth latent structure. Simulated expression matrices were generated from a nonnegative factor model in which prognostic gene programs explained low variance relative to outcome-neutral background signals, with survival times generated from latent factor scores via a Cox model ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Methods). We compared DeSurv to standard NMF ($\alpha=0$) followed by post hoc Cox regression, with both methods tuned via cross-validated concordance index using Bayesian optimization.

In the primary scenario, where prognostic programs explained low variance relative to outcome-neutral programs, DeSurv consistently achieved higher C-index and substantially improved precision---the fraction of genes in a learned factor that belong to a true prognostic gene program---for recovering the true prognostic gene programs (Fig. \ref{fig:sim}A-B). The unsupervised baseline exhibited near-zero precision, indicating that variance-driven factorization failed to concentrate on the genes that actually drove survival. This confirms that outcome-guided learning recovers prognostic programs more reliably when variance and prognosis diverge. Across 100 replicates, DeSurv achieved median C-index $`r desurv_sim_median`$ versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$).

To verify that these gains reflect genuine signal recovery rather than overfitting, we repeated the analysis under two additional simulation scenarios (SI Appendix, Fig. S2). In a null scenario where survival times were generated independently of gene expression ($\beta = 0$), DeSurv defaulted to standard NMF: Bayesian optimization selected low supervision strength ($\alpha$), and both methods yielded C-index values near 0.5. This indicates that the semi-supervised design does not impose spurious prognostic structure when none exists. In a mixed scenario where survival depended on both factor-specific marker genes and shared background genes, creating partial overlap between variance-dominant and prognostically relevant structure, DeSurv's advantage was present but attenuated ($\Delta$ median C-index $= `r delta_mixed_cindex`$) relative to the primary scenario ($\Delta = `r delta_sim_cindex`$). Together, these three scenarios establish that DeSurv's benefit scales with the degree of divergence between variance and prognosis: largest when prognostic programs explain low variance (primary scenario), moderate when variance and prognosis partially overlap (mixed), and absent when no survival signal exists (null).

```{r fig-sim-ntop-scenarios, fig.width = 6.5, fig.height = 3, fig.cap = "Performance comparison between DeSurv and standard NMF ($\\alpha = 0$) in the primary simulation scenario ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$, 100 replicates), where prognostic programs explain low variance relative to outcome-neutral programs. Both methods were tuned via Bayesian optimization over cross-validated concordance index. (A) Test-set concordance index across simulation replicates. (B) Precision (fraction of genes in a learned factor belonging to a true prognostic gene program) across simulation replicates. Each point represents one replicate. \\label{fig:sim}", fig.env='figure*', fig.pos='t', out.width='\\textwidth'}



plot_grid(alt_plots$cindex_box + labs(title=NULL) + scale_y_continuous(limits=c(.5,1)),
          alt_plots$precision_box + labs(title=NULL),ncol=2,labels=c("A","B"))
```



## Survival supervision reorganizes the learned factor structure in PDAC {.unnumbered}

To directly test whether survival supervision reorganizes the learned factor structure, as predicted by the simulation results above, we examined the overlap between factor-specific gene rankings and established PDAC gene programs (Fig. \ref{fig:bio}A-B). Bayesian optimization selected $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$ and $\alpha = `r round(tar_read(tar_params_best_tcgacptac)$alpha, 3)`$ for DeSurv in the PDAC training cohorts (TCGA and CPTAC). To enable a direct comparison of how each method organizes the same number of factors, we also fit standard NMF at the same rank ($k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$); results for standard NMF at independently selected ranks ($k = 5$ via the residual elbow heuristic, $k = 7$ via cross-validated concordance at $\alpha = 0$) are presented in the SI Appendix. We note that standard NMF at $k = 5$ achieves comparable or slightly higher external validation concordance relative to DeSurv at $k = 3$; the DeSurv advantage lies in parsimony — recovering equivalent prognostic generalization with fewer, more interpretable factors — rather than in raw concordance at each method's unrestricted optimum.

We refer to the three DeSurv factors as D1--D3 and the three NMF factors as N1--N3. DeSurv produced a factorization that concentrated survival signal into a single factor (D1) integrating classical tumor programs with an iCAF-associated stromal signature (Fig. \ref{fig:bio}A). This is consistent with evidence that survival stratification in PDAC improves when tumor-intrinsic and CAF subtypes are considered jointly [@peng2024determination; @elyada2019cross]. A second factor (D2) captured an activated immune-stromal microenvironment. A third factor (D3) was enriched for basal-like tumor programs but, as discussed below, contributed minimally to survival in conditional analysis. This reflects two features of the factorization: D3 absorbs residual exocrine-compositional variation alongside basal-like programs, and the prognostically relevant tumor-stroma contrast is already concentrated within D1. Notably, no DeSurv factor showed a significant positive correlation with exocrine-associated gene programs (DECODER Exocrine, Collisson Exocrine, Bailey ADEX; Fig. \ref{fig:bio}A), suggesting that survival supervision deprioritizes differentiation-related variation that is weakly associated with outcome.

At the same rank, standard NMF organized the three factors around dominant sources of transcriptional variance rather than prognostic biology, as expected for an unsupervised method optimizing reconstruction fidelity (Fig. \ref{fig:bio}B). One factor (N2) was strongly and significantly correlated with exocrine-associated gene programs (DECODER Exocrine, Collisson Exocrine, Bailey ADEX; Fig. \ref{fig:bio}B) and negatively correlated with immune and stromal signatures, consistent with a bulk composition axis separating normal or differentiated tissue from non-epithelial tumor content. A second factor (N1) aligned with classical tumor identity, while a third (N3) aggregated immune, fibroblast, and extracellular matrix-related programs — including Puleo Desmoplastic, SCISSORS panCAF and myCAF, Maurer Immune and ECM, and DECODER Activated Stroma — into a single composite microenvironmental factor (Fig. \ref{fig:bio}B). These patterns indicate that, given the same number of factors, standard NMF allocates substantial model capacity to differentiation-driven signals and merges distinct microenvironmental states that DeSurv separates.

We quantified these differences by contrasting the conditional fraction of expression variance explained by each factor with its survival contribution (SI Appendix, Fig. S7). A striking pattern emerged: all three NMF factors contributed negligibly to survival despite spanning a range of conditional variance, while DeSurv concentrated nearly all survival signal into D1. Specifically, the DeSurv factor with the largest survival contribution explained $`r desurv_maxsurv_pct`$% of variance with $\Delta\ell = `r desurv_maxsurv_dl`$, while the highest-variance NMF factor explained $`r nmf_maxvar_pct`$% of expression variance but contributed only $\Delta\ell = `r nmf_maxvar_dl`$ to survival. Notably, the highest-variance DeSurv factor (D3, ${\sim}30\%$ of conditional variance) also contributed minimally to survival after accounting for D1. This apparent paradox — given the well-established prognostic role of basal-like biology in PDAC [@moffitt2015virtual;@rashid2020purity] — reflects two features of the factorization. First, D3 absorbs residual exocrine-compositional variation alongside basal-like programs (see factor correspondence below), diluting its prognostic specificity. Second, the survival-relevant contrast between classical and basal-like biology is already captured by D1's iCAF-classical co-program, leaving D3's conditional contribution minimal. This illustrates the misalignment between variance and prognosis anticipated by sufficient dimension reduction theory [@cook2007fisher] and observed empirically in tumor purity analyses [@aran2015systematic]: the factors that explain the most transcriptomic variation need not be the factors most associated with patient outcomes. The observed pattern in PDAC is consistent with the intermediate regime between the primary and mixed simulation scenarios, where DeSurv's advantage was largest.

To further characterize how the learned factor structure differs between DeSurv and standard NMF, we examined the W-matrix gene loading correspondence between factors derived from the two methods (SI Appendix, Fig. S8). NMF's classical-associated factor (N1) showed the strongest correspondence with DeSurv's basal-like factor (D3), and NMF's composite microenvironmental factor (N3) mapped primarily to DeSurv's microenvironmental factor (D2). The exocrine-like NMF factor (N2) did not correspond strongly to any single DeSurv factor; its signal was weakly absorbed into D3, consistent with D3's high conditional variance and low survival contribution described above. Together, these results demonstrate that DeSurv selectively preserves biologically distinct tumor and microenvironmental structure while reorganizing variance-dominant but prognostically inert variation — including exocrine-compositional signal — into a high-variance, low-survival factor rather than allocating dedicated model capacity to it.
```{r fig-bio, fig.cap="DeSurv reorganizes transcriptional structure and generalizes to held-out cohorts. (A--B) Spearman correlations between factor gene rankings and established PDAC gene programs for DeSurv (A) and standard NMF (B). Asterisks denote significance after multiple testing correction; only correlations $> 0.2$ shown. (C--D) Kaplan--Meier curves for pooled external validation samples stratified by the cross-validated optimal cutpoint for the DeSurv linear predictor (C) and the analogous NMF-derived factor (D). \\label{fig:bio}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

library(cowplot)
library(ggplot2)

tar_load(fig_gene_overlap_heatmap_desurv_tcgacptac)
tar_load(fig_gene_overlap_heatmap_std_desurvk_tcgacptac)
tar_load(fig_median_survival_desurv_tcgacptac)
tar_load(fig_median_survival_std_desurvk_tcgacptac)

# ── Extract dichotomized KM HR/CI/p from figure annotation ───────────────────
km_extract_label <- function(splot_obj) {
  for (layer in splot_obj$plot$layers) {
    # annotate() stores fixed aesthetics in aes_params in most ggplot2 versions
    lbl <- tryCatch(layer$aes_params$label, error = function(e) NULL)
    if (!is.null(lbl) && length(lbl) > 0 && grepl("HR", lbl[1], fixed = TRUE)) return(lbl[1])
    # Fallback: try data$label (some ggplot2 versions store it here)
    lbl <- tryCatch({
      d <- layer$data
      if (is.function(d)) d <- tryCatch(d(data.frame()), error = function(e) NULL)
      if (is.data.frame(d) && "label" %in% names(d)) d$label[1] else NULL
    }, error = function(e) NULL)
    if (!is.null(lbl) && length(lbl) > 0 && grepl("HR", lbl[1], fixed = TRUE)) return(lbl[1])
  }
  NULL
}
km_parse_stats <- function(lbl) {
  if (is.null(lbl)) return(list(hr = NA, ci_lo = NA, ci_hi = NA, p_lt = TRUE, p_val = NA))
  nums  <- as.numeric(regmatches(lbl, gregexpr("[0-9]+\\.[0-9]+", lbl))[[1]])
  p_lt  <- grepl("< 0.001", lbl, fixed = TRUE)
  p_val <- if (!p_lt) as.numeric(sub(".*p = ([0-9.]+).*", "\\1", lbl)) else NA_real_
  list(hr = nums[1], ci_lo = nums[2], ci_hi = nums[3], p_lt = p_lt, p_val = p_val)
}
km_desurv <- km_parse_stats(km_extract_label(fig_median_survival_desurv_tcgacptac))
km_nmf    <- km_parse_stats(km_extract_label(fig_median_survival_std_desurvk_tcgacptac))

# -------------------------
# Global styling
# -------------------------
base_size <- 8
theme_pnas <- theme_classic(base_size = base_size) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.margin = margin(6, 10, 6, 15),
    legend.box.margin = margin(0, 0, 0, 0),
    legend.spacing.x = unit(6, "pt"),
    legend.key.width = unit(12, "pt")
  )

# -------------------------
# Helper: stack KM plot + risk table
# -------------------------
km_text_size        <- base_size - 3   # axis labels and tick numbers in KM panels
risk_title_size     <- 5               # "Number at risk" label

stack_surv <- function(surv_obj, title,keep_legend = FALSE) {
  p <- surv_obj$plot +
    theme_pnas +
    theme(
      legend.position = if (keep_legend) "bottom" else "none",
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = km_text_size),
      axis.text.x  = element_text(size = km_text_size),
      axis.text.y  = element_text(size = km_text_size),
      plot.margin = margin(10, 10, 2, 10),
      plot.title = element_text(size = risk_title_size+3)
    )+
    labs(title=title)
  
  t <- surv_obj$table +
    theme_pnas +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.x  = element_text(size = km_text_size),
      axis.text.y  = element_text(size = km_text_size),
      text = element_text(size = km_text_size),
      axis.title.x = element_text(size = km_text_size),
      plot.title   = element_text(size = risk_title_size),
      plot.margin  = margin(10, 10, 6, 10)
    ) +
    labs(x = "Time (months)")
    t$layers[[1]]$aes_params$size = km_text_size / ggplot2::.pt
  plot_grid(p, t, ncol = 1, rel_heights = c(2.5, 1.5), align = "v", axis = "lr")
}

# -------------------------
# Panels A and B (heatmaps) - side by side on left
# -------------------------
plot_4a    <- fig_gene_overlap_heatmap_desurv_tcgacptac$plot
plot_4b    <- fig_gene_overlap_heatmap_std_desurvk_tcgacptac$plot
legend_4ab <- fig_gene_overlap_heatmap_desurv_tcgacptac$legend
legend_4ab <- gtable::gtable_add_padding(
  legend_4ab,
  padding = unit(c(0, 10, 0, 0), "pt")  # top, right, bottom, left
)

left_block <- plot_grid(
  plot_4a, plot_4b, cowplot::ggdraw(legend_4ab),
  ncol = 3, labels = c("A", "B", ""),align="hv",
  label_size = 12, 
  rel_widths = c(2.5, 2.5, 0.6)
)

# -------------------------
# Panels C and D (KM curves) - stacked vertically on right
# -------------------------
med_desurv <- fig_median_survival_desurv_tcgacptac
med_std    <- fig_median_survival_std_desurvk_tcgacptac

km_c <- stack_surv(med_desurv, "DeSurv", keep_legend = FALSE)
km_d <- stack_surv(med_std, "NMF",   keep_legend = FALSE)

legend_plot =ggplot(
    data.frame(x = 1:2, y = 1:2, group = factor(c("Low", "High"), levels = c("Low", "High"))),
    aes(x = x, y = y, colour = group)
  ) +
    geom_line() +
    scale_colour_manual(values = c("Low" = "violetred2", "High" = "turquoise4"),
                        name = "Risk group") +
    theme_pnas +
    theme(legend.position = "bottom",legend.text = element_text(size=km_text_size),
          legend.title = element_text(size=km_text_size)) +
    guides(colour = guide_legend(nrow = 1))

legend_grob = ggplotGrob(legend_plot)

legend = legend_grob$grobs[
    sapply(legend_grob$grobs, function(x) x$name) == "guide-box"
][[1]]


right_block <- plot_grid(
  km_c, km_d, ggdraw(legend),
  nrow = 3,
  labels = c("C", "D", ""),
  label_size = 12,
  rel_heights = c(5, 5, 1)
)

# -------------------------
# Full figure: heatmaps (left) + KM curves stacked (right)
# -------------------------
plot_grid(left_block, right_block, ncol = 2, rel_widths = c(6, 3)) 
```


## Survival-aligned programs generalize across independent PDAC cohorts {.unnumbered}

Having shown that DeSurv reorganizes the learned factor structure in the training data, we next tested whether this structure generalizes to independent cohorts. If outcome-guided subspaces capture biologically reproducible rather than noise-driven structure, they should transfer more readily across datasets [@bair2006supervised]. We computed factor scores in each validation dataset by projecting the gene programs learned in the training cohort ($Z = \tilde{W}^\top X_{\text{new}}$), yielding a continuous score for each factor per sample.

For each method, both fit at $k = 3$ (the DeSurv BO-selected rank), we focused on the factor showing the largest increase in Cox model log partial likelihood in the training data. External validation C-index for NMF at independently selected ranks is summarized in SI Appendix, Table S4. Across `r n_val_cohorts` independent PDAC cohorts ($n = `r n_val_total`$), the DeSurv-derived factor exhibited consistent survival effects, with hazard ratio estimates showing limited variability and predominantly protective associations (continuous risk score model: pooled HR $`r hr_pool`$; 95% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$; $P = `r p_pool`$; SI Appendix, Fig. S9). In contrast, the NMF factor identified by the same criterion showed greater heterogeneity across datasets and weaker survival associations, suggesting that variance-driven programs may capture cohort-specific variation that does not transfer as reliably.

To assess clinical interpretability of these continuous associations, we dichotomized validation samples into high- and low-risk groups using independently optimized cross-validated cutpoints for each method. The DeSurv linear predictor produced clear separation of survival trajectories (pooled validation HR $= `r km_desurv$hr`$, 95% CI $`r km_desurv$ci_lo`$--$`r km_desurv$ci_hi`$, $P `r if(km_desurv$p_lt) "< 0.001" else paste0("= ", km_desurv$p_val)`$; Fig. \ref{fig:bio}C). The same procedure applied to NMF yielded weaker stratification (HR $= `r km_nmf$hr`$, 95% CI $`r km_nmf$ci_lo`$--$`r km_nmf$ci_hi`$, $P `r if(km_nmf$p_lt) "< 0.001" else paste0("= ", km_nmf$p_val)`$; Fig. \ref{fig:bio}D).


## Factorization rank K=3 is robustly confirmed across supervision strength and gene-projection settings {.unnumbered}

```{r ksens-counts, echo=FALSE}
# ── Load K-sensitivity RDS outputs ────────────────────────────────────────────
# Generated by:
#   Rscript inst/cv_grid_validation_analysis_amber_optimal.R  → adj_p_270, hcor_270, master_rows_270
#   Rscript inst/cv_grid_validation_analysis_allgenes.R       → adj_p_all, hcor_all, master_rows_all
#   Rscript inst/k3_k7_nesting_heatmap.R                     → k3_k7_summary
#   Rscript inst/create_ksens_rds.R                          → lam300_summary, production_summary (bootstrap)
RDS_DIR <- "results/cv_grid"

adj_p_270  <- readRDS(file.path(RDS_DIR, "adj_p_270_matrix.rds"))
adj_p_all  <- readRDS(file.path(RDS_DIR, "adj_p_all_matrix.rds"))
hcor_270   <- readRDS(file.path(RDS_DIR, "hcor_270_matrix.rds"))
hcor_all   <- readRDS(file.path(RDS_DIR, "hcor_all_matrix.rds"))
master_270 <- readRDS(file.path(RDS_DIR, "master_rows_270.rds"))
master_all <- readRDS(file.path(RDS_DIR, "master_rows_all.rds"))
lam300     <- readRDS(file.path(RDS_DIR, "lam300_summary.rds"))
prod_sum   <- readRDS(file.path(RDS_DIR, "production_summary.rds"))

# ── Derived scalars from p-value matrices ─────────────────────────────────────
n_sig_270        <- sum(adj_p_270 < 0.05)
n_sig_all        <- sum(adj_p_all < 0.05)
n_k3_sig_270     <- sum(adj_p_270["K=3", ] < 0.05)
k3_alpha55_p_270 <- adj_p_270["K=3", "a=0.55"]
k3_alpha55_p_all <- adj_p_all["K=3", "a=0.55"]
k5_alpha55_p_270 <- adj_p_270["K=5", "a=0.55"]
k5_alpha25_p_all <- adj_p_all["K=5", "a=0.25"]
k7_alpha75_p_270 <- adj_p_270["K=7", "a=0.75"]

# ── H-cor values from hcor matrices ───────────────────────────────────────────
hcor_k3_a55_270  <- hcor_270["K=3", "a=0.55"]
hcor_k5_a55_270  <- hcor_270["K=5", "a=0.55"]
hcor_k5_a25_all  <- hcor_all["K=5", "a=0.25"]

# ── λ=0.300 analysis scalars ───────────────────────────────────────────────────
n_sig_all_prior  <- lam300$n_sig
production_adj_p <- prod_sum$adj_p

# ── Grid dimensions (for "of N combinations" and "of M alpha values" prose) ──
GRID_K      <- c(2, 3, 5, 7, 9)
GRID_ALPHA  <- c(0, 0.25, 0.35, 0.55, 0.75, 0.85, 0.95)
n_alpha     <- length(GRID_ALPHA)
n_grid      <- length(GRID_K) * length(GRID_ALPHA)
```

The 1-SE model selection rule chose $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$ over the global Bayesian optimization maximum ($k = `r tar_read(tar_k_selection_tcgacptac)$k_best`$; CV C-index $`r k_best_cindex`$ versus $`r k3_best_cindex`$ for $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$; margin $`r k_margin`$ C-index units). Without this rule, DeSurv would have selected $k = `r tar_read(tar_k_selection_tcgacptac)$k_best`$ at $\alpha = `r round(tar_read(tar_params_best_tcgacptac)$alpha, 3)`$---the same rank independently favored by standard NMF under cross-validated concordance at $\alpha = 0$. The parsimony of the DeSurv solution relative to standard NMF therefore reflects a deliberate methodological choice encoded in the 1-SE rule, not an intrinsic property of survival supervision. Whether this preference is empirically warranted requires evaluating whether the additional complexity of $k = `r tar_read(tar_k_selection_tcgacptac)$k_best`$ translates to improved generalization.

Bayesian optimization selected $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$ and $\alpha = `r round(tar_read(tar_params_best_tcgacptac)$alpha, 3)`$ as the primary model configuration, but whether this reflects a stable solution or a fragile local optimum required further evaluation. To assess robustness, we conducted an exhaustive sensitivity analysis across $K \in \{2, 3, 5, 7, 9\}$ and $\alpha \in \{0, 0.25, 0.35, 0.55, 0.75, 0.85, 0.95\}$ (`r n_grid` combinations), holding all remaining hyperparameters fixed at the production-optimal values ($\lambda = 0.349$, $\xi = 0.056$, $n_{\text{top}} = 270$, 100 initializations). For each $(K, \alpha)$ combination, we identified the factor most correlated with the production $K = 3$ Factor 1 and assessed its survival association in `r prod_sum$n_val_total` independent validation patients across `r prod_sum$n_val_cohorts` external cohorts, using stratified Cox regression with and without adjustment for PurIST and DeCAF classifiers. (PACA-AU array and RNA-seq are treated as a single cohort in this sensitivity analysis; the generalization analysis above counts them separately, yielding $n = `r n_val_total`$ across `r n_val_cohorts` datasets.)

With the production-optimal gene-projection setting ($n_{\text{top}} = 270$), $K = 3$ achieved adjusted significance at `r n_k3_sig_270` of `r n_alpha` supervision-strength values, including $\alpha = 0.55$ where it achieved adjusted HR $= `r round(master_270[master_270$nm == "K3_a055", "hr_v_adj"], 2)`$ ($p = `r formatC(k3_alpha55_p_270, format="f", digits=3)`$) --- nearly identical to the production fit ($p = `r formatC(production_adj_p, format="f", digits=3)`$). $K = 5$ at $\alpha = 0.55$ provided independent corroboration at higher factor-program fidelity (Spearman H-score correlation = `r round(hcor_k5_a55_270, 3)` vs. $K = 3$'s `r round(hcor_k3_a55_270, 3)`; adjusted $p = `r formatC(k5_alpha55_p_270, format="f", digits=3)`$), confirming that the iCAF biology is recoverable by a higher-rank model. In contrast, $K = 7$ --- which had appeared to achieve adjusted significance ($p = `r formatC(lam300$best_adj_p, format="f", digits=3)`$) in a parallel analysis using all `r prod_sum$n_genes_total` genes in the projection --- did not achieve adjusted significance with $n_{\text{top}} = 270$ (adjusted $p = `r formatC(k7_alpha75_p_270, format="f", digits=3)`$), indicating that the earlier $K = 7$ advantage depended on the all-genes projection rather than reflecting an intrinsic property of the 7-factor solution.

To isolate the contribution of gene focusing, we repeated the analysis with all genes retained ($n_{\text{top}} = \text{NULL}$) while keeping $\lambda$ and $\xi$ at their optimal values. With all genes, $K = 3$ at $\alpha = 0.55$ reached only adjusted $p = `r formatC(k3_alpha55_p_all, format="f", digits=3)`$, failing to reproduce the production result, and only `r n_sig_all` of `r n_grid` combinations achieved adjusted $p < 0.05$ --- compared with `r n_sig_270` of `r n_grid` with $n_{\text{top}} = 270$. Substituting $\lambda = 0.349$ for $\lambda = 0.300$ (the original and BO-optimized values) within the all-genes setting shifted the pattern of validated combinations but left the overall count unchanged (`r n_sig_all_prior`/`r n_grid`), indicating that regularization tuning alone cannot reproduce the adjusted signal without gene focusing. These analyses demonstrate that $n_{\text{top}} = 270$ is the critical hyperparameter: focusing the projection onto the 270 most factor-characteristic genes concentrates the survival signal, suppresses noise from the remaining genes, and prevents overfitting at high supervision weights. Full $K \times \alpha$ adjusted p-value tables are provided in the SI Appendix (Tables S1--S3). Taken together, these results provide empirical justification for the 1-SE rule's parsimony preference: $k = `r tar_read(tar_k_selection_tcgacptac)$k_best`$, which achieved the highest training CV C-index, did not achieve adjusted external validation at the production gene-focusing setting ($p = `r formatC(k7_alpha75_p_270, format="f", digits=3)`$ with $n_{\text{top}} = 270$), while $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$ validated robustly ($p = `r formatC(k3_alpha55_p_270, format="f", digits=3)`$). The simpler model not only passed the 1-SE threshold but also generalized more reliably.
