<!-- REVISED RESULTS — prediction-validation structure for PNAS -->
<!-- All R code chunks are preserved from 04_results.Rmd -->
<!-- Prose revisions: bolded one-sentence summaries, prediction-validation transitions -->

<!-- SI Appendix figure numbering (supplement.Rmd order):
     S1  fig-converge       convergence trajectories
     S2  fig-sim-null-mixed null and mixed simulation scenarios
     S3  fig-cindex-by-k    C-index vs K sensitivity
     S4  fig-k3-k7-nesting  K=3 vs K=7 factor correspondence
     S5  fig-cutpoint-km    cutpoint selection and KM curves
     S6  fig-subtype-overlap subtype composition of risk groups
     S7  fig-varsurvival    variance explained vs survival contribution
     S8  fig-corr           NMF vs DeSurv factor W-matrix correspondence
     S9  fig-forest         per-cohort hazard ratio forest plot
-->

```{r setup}
library(targets)
library(stringr)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(cowplot)
library(enrichplot)
library(magick)
library(Seurat)
library(gt)
library(VAM)
library(grid)
library(viridis)
library(ggplotify)
library(NMF)


tar_load(sim_figs_by_scenario)

scenario_ids = sapply(sim_figs_by_scenario,function(x) x$scenario_id)
analysis_ids = sapply(sim_figs_by_scenario,function(x) x$analysis_id)

alt = which(scenario_ids=="R0_easy" & analysis_ids == "bo_tune_ntop")
null = which(scenario_ids=="R00_null" & analysis_ids == "bo_tune_ntop")

alt_plots = sim_figs_by_scenario[[alt]]
null_plots = sim_figs_by_scenario[[null]]

alt_plots = lapply(alt_plots, set_fig_font, size = 10)
null_plots = lapply(null_plots, set_fig_font, size = 10)

mixed = which(scenario_ids=="R_mixed" & analysis_ids == "bo_tune_ntop")
mixed_plots = sim_figs_by_scenario[[mixed]]
mixed_plots = lapply(mixed_plots, set_fig_font, size = 10)
```

```{r computed-stats, echo=FALSE, message=FALSE, warning=FALSE}
# ── Simulation C-index statistics (from alt_plots already loaded above) ──────
cindex_dat        <- alt_plots$cindex_box$data
desurv_sim_median <- round(median(cindex_dat$cindex[cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_sim_median    <- round(median(cindex_dat$cindex[cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_sim_cindex  <- round(desurv_sim_median - nmf_sim_median, 3)

# ── Paired Wilcoxon test (matched replicates) ────────────────────────────────
# Each replicate uses the same simulated dataset for both methods
sim_wide <- merge(
  cindex_dat[cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  cindex_dat[cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
sim_wilcox_p <- signif(wilcox.test(sim_wide$cindex_desurv, sim_wide$cindex_nmf,
                                    paired = TRUE)$p.value, 2)

# ── Mixed-scenario C-index statistics ────────────────────────────────────────
mixed_cindex_dat        <- mixed_plots$cindex_box$data
desurv_mixed_median <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "DeSurv"], na.rm = TRUE), 3)
nmf_mixed_median    <- round(median(mixed_cindex_dat$cindex[mixed_cindex_dat$method == "NMF"],    na.rm = TRUE), 3)
delta_mixed_cindex  <- round(desurv_mixed_median - nmf_mixed_median, 3)

# ── Paired Wilcoxon test (mixed scenario) ────────────────────────────────────
mixed_wide <- merge(
  mixed_cindex_dat[mixed_cindex_dat$method == "DeSurv", c("replicate", "cindex")],
  mixed_cindex_dat[mixed_cindex_dat$method == "NMF",    c("replicate", "cindex")],
  by = "replicate", suffixes = c("_desurv", "_nmf")
)
mixed_wilcox_p <- signif(wilcox.test(mixed_wide$cindex_desurv, mixed_wide$cindex_nmf,
                                      paired = TRUE)$p.value, 2)

# ── Variance vs survival statistics for PDAC training data ───────────────────
tar_load(fig_variation_explained_tcgacptac)
var_surv_dat   <- fig_variation_explained_tcgacptac$data
nmf_vs         <- var_surv_dat[var_surv_dat$method == "NMF", ]
desurv_vs      <- var_surv_dat[var_surv_dat$method == "DeSurv", ]
nmf_maxvar     <- nmf_vs[which.max(nmf_vs$variance_explained), ]
desurv_maxsurv <- desurv_vs[which.max(desurv_vs$delta_loglik), ]
nmf_maxvar_pct      <- round(nmf_maxvar$variance_explained * 100, 1)
nmf_maxvar_dl       <- round(nmf_maxvar$delta_loglik, 1)
desurv_maxsurv_pct  <- round(desurv_maxsurv$variance_explained * 100, 1)
desurv_maxsurv_dl   <- round(desurv_maxsurv$delta_loglik, 1)
desurv_maxvar       <- desurv_vs[which.max(desurv_vs$variance_explained), ]
desurv_maxvar_pct   <- round(desurv_maxvar$variance_explained * 100, 1)

# ── Pooled PDAC validation statistics ────────────────────────────────────────
tar_load(val_latent_desurv_tcgacptac)
pool_dfs <- lapply(val_latent_desurv_tcgacptac, function(entry) {
  surv <- entry$survival
  if (is.null(surv) || !all(c("time", "event") %in% names(surv))) return(NULL)
  data.frame(time = surv$time, event = surv$event,
             dataset = entry$dataset, risk_score = entry$risk_score)
})
pool_df      <- dplyr::bind_rows(Filter(Negate(is.null), pool_dfs))
pool_df      <- pool_df[is.finite(pool_df$time) & is.finite(pool_df$risk_score), ]
n_val_total  <- nrow(pool_df)
n_val_events <- sum(pool_df$event, na.rm = TRUE)
n_val_cohorts <- length(unique(pool_df$dataset))
cox_pool     <- coxph(Surv(time, event) ~ risk_score + strata(dataset), data = pool_df)
cox_pool_sum <- summary(cox_pool)
hr_pool      <- round(cox_pool_sum$coefficients[, "exp(coef)"], 2)
ci_pool_lo   <- round(cox_pool_sum$conf.int[, "lower .95"], 2)
ci_pool_hi   <- round(cox_pool_sum$conf.int[, "upper .95"], 2)
p_pool       <- signif(cox_pool_sum$coefficients[, "Pr(>|z|)"], 2)

# ── NMF pooled continuous HR (generated by inst/compute_nmf_pooled_hr.R) ─────
nmf_pool_rds <- file.path("..", "results", "nmf_pooled_hr.rds")
if (file.exists(nmf_pool_rds)) {
  nmf_pool     <- readRDS(nmf_pool_rds)
  hr_pool_nmf  <- nmf_pool$hr
  ci_pool_lo_nmf <- nmf_pool$ci_lo
  ci_pool_hi_nmf <- nmf_pool$ci_hi
  p_pool_nmf   <- nmf_pool$p
} else {
  warning("results/nmf_pooled_hr.rds not found. Run: Rscript inst/compute_nmf_pooled_hr.R")
  hr_pool_nmf <- ci_pool_lo_nmf <- ci_pool_hi_nmf <- p_pool_nmf <- "NA"
}

# ── K-selection C-index values from BO history ───────────────────────────────
k_sel          <- tar_read(tar_k_selection_tcgacptac)
k_best_cindex  <- round(k_sel$best_mean, 3)
bo_hist        <- tar_read(desurv_bo_results_tcgacptac)$history
bo_hist        <- bo_hist[!is.na(bo_hist$mean_cindex) & bo_hist$status == "ok", ]
k3_hist        <- bo_hist[bo_hist$k_grid == k_sel$k_selected, ]
k3_best_cindex <- if (nrow(k3_hist)) round(max(k3_hist$mean_cindex), 3) else NA_real_
k_margin       <- round(k_best_cindex - k3_best_cindex, 4)
```

## Model Overview {.unnumbered}

We developed DeSurv, a survival-supervised deconvolution framework (Fig. \ref{fig:schema}; Methods), and evaluated it in controlled simulations and PDAC cohort analysis with external validation.
```{r fig-schema, fig.cap = "Overview of the DeSurv framework. (A) DeSurv jointly optimizes NMF reconstruction ($X \\approx WH$) and Cox proportional hazards likelihood. The gene program matrix $W$ is shared between objectives: factor scores $Z = W^\\top X$ serve as covariates in the Cox model with coefficients $\\beta$. A supervision parameter $\\alpha \\in [0,1)$ controls the balance between reconstruction fidelity ($\\alpha = 0$, standard NMF) and survival prediction ($\\alpha$ close to 1). (B) The factorization rank ($k$), supervision strength ($\\alpha$), and regularization ($\\lambda$) are selected via Bayesian optimization over cross-validated concordance index. \\label{fig:schema}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width='\\textwidth'}

knitr::include_graphics(file.path("..","figures", "model_schematic_final.pdf"))
```


## Survival supervision clarifies NMF rank selection {.unnumbered}

We first asked whether incorporating survival outcomes resolves the instability of unsupervised rank selection heuristics [@frigyesi2008nmf; @brunet2004metagenes]. Using gene expression data from the PDAC training cohorts (TCGA [@raphael2017integrated] and CPTAC [@ellis2013clinical]; Methods), we evaluated commonly used unsupervised criteria across a range of candidate ranks.

Standard NMF diagnostics yielded inconsistent guidance (Fig. \ref{fig:bo}A-C). Reconstruction residuals decreased smoothly with increasing $k$ and did not exhibit a clear elbow, a pattern consistent with both relatively small solutions ($k \approx 3$-4) and substantially larger ranks ($k \approx 6$-8). The cophenetic correlation coefficient began to decline at low ranks ($k \approx 3$-4) but continued to fluctuate at higher values without a distinct transition point. In contrast, mean silhouette width was highest at very small ranks ($k \approx 2$-3) and decreased monotonically thereafter, favoring low-dimensional solutions that conflicted with the other criteria. Together, these unsupervised heuristics pointed to incompatible values of $k$, illustrating the ambiguity of rank selection in standard NMF.

Because the goal is prognostic factorization, DeSurv selects $k$ using the criterion that directly measures prognostic performance: cross-validated concordance index (C-index), optimized jointly with supervision strength $\alpha$. The resulting C-index surface identified a coherent region of high performance (Fig. \ref{fig:bo}D). Model selection followed the one-standard-error rule (Methods): we selected the smallest $k$ whose predicted performance lay within one standard error of the maximum, yielding a parsimonious choice. Applying the 1-SE rule to the Bayesian optimization surface selected $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$ and $\alpha = `r round(tar_read(tar_params_best_tcgacptac)$alpha, 3)`$ (peak C-index $`r round(tar_read(tar_k_selection_tcgacptac)$best_mean, 3)`$; $n = `r ncol(tar_read(tar_data_filtered_tcgacptac)$ex)`$ patients, $`r sum(tar_read(tar_data_filtered_tcgacptac)$sampInfo$event)`$ events). Survival-supervised model selection thus resolves the rank ambiguity that unsupervised heuristics leave open, yielding a single parsimonious factorization for downstream evaluation.

```{r fig-sim, fig.width = 6.5, fig.height = 6, fig.cap = "Simulation studies and model selection. (A) Test-set concordance index across 100 simulation replicates ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$) comparing DeSurv to standard NMF ($\\alpha = 0$). (B) Precision (fraction of top-weighted genes in a learned factor overlapping a true prognostic program) across replicates. (C) Distribution of selected $k$ values for DeSurv versus standard NMF. (D) Gaussian process predicted mean cross-validated C-index from Bayesian optimization over factorization rank ($k$) and supervision strength ($\\alpha$) in PDAC training data (TCGA + CPTAC). \\label{fig:sim}", fig.env='figure*', fig.pos='t', out.height= "6in", out.width="6.5in"}

tar_load(fig_bo_heat_tcgacptac)
fig_bo_heat_tcgacptac = set_fig_font(fig_bo_heat_tcgacptac, size = 10)

upper = plot_grid(
  alt_plots$cindex_box + labs(title = NULL) + scale_y_continuous(limits = c(.5, 1)),
  alt_plots$precision_box + labs(title = NULL),
  ncol = 2, labels = c("A", "B")
)

lower = plot_grid(
  alt_plots$k_hist,
  fig_bo_heat_tcgacptac,
  ncol = 2, labels = c("C", "D")
)

plot_grid(upper, lower, nrow = 2)

```


## DeSurv recovers prognostic gene programs when variance and prognosis diverge {.unnumbered}

To test whether survival supervision improves gene program recovery under nonnegative constraints, we designed simulations with known ground-truth latent structure. Simulated expression matrices were generated from a nonnegative factor model in which prognostic gene programs explained low variance relative to outcome-neutral background signals, with survival times generated from latent factor scores via a Cox model ($p = 3{,}000$ genes, $n = 200$ samples, true $k = 3$; Methods). To test whether survival information should enter during factorization rather than only afterward, we compared DeSurv to standard NMF ($\alpha=0$), where factors are learned by reconstruction error alone and survival associations are estimated by a separately fit Cox model. Both methods were tuned via Bayesian optimization over cross-validated C-index, so that differences in performance reflect the role of supervision during factorization, not during model selection.

In the primary scenario, where prognostic programs explained low variance relative to outcome-neutral programs, DeSurv consistently achieved higher C-index (Fig. \ref{fig:sim}A) and substantially higher precision for recovering the true prognostic gene programs (Fig. \ref{fig:sim}B), where precision denotes the fraction of top-weighted genes in a learned factor that overlap with a true prognostic program's gene set. The unsupervised baseline exhibited near-zero precision, indicating that variance-driven factorization failed to concentrate on the genes that actually drove survival. This confirms that outcome supervision during factorization is necessary to recover prognostic programs that do not dominate expression variance. Across 100 replicates, DeSurv achieved median C-index $`r desurv_sim_median`$ versus $`r nmf_sim_median`$ for standard NMF ($\Delta = `r delta_sim_cindex`$, paired Wilcoxon $P = `r sim_wilcox_p`$).

To verify that these gains reflect genuine signal recovery rather than overfitting, we repeated the analysis under a null scenario ($\beta = 0$) and a mixed scenario with partial overlap between variance-dominant and prognostic structure (SI Appendix, Fig. S2). DeSurv's benefit scaled with the degree of divergence between variance and prognosis: largest in the primary scenario ($\Delta = `r delta_sim_cindex`$), attenuated when variance and prognosis partially overlapped ($\Delta = `r delta_mixed_cindex`$, paired Wilcoxon $P = `r mixed_wilcox_p`$), and absent when no survival signal existed, where both methods yielded C-index near 0.5. DeSurv also recovered the true rank ($k = 3$) more frequently than standard NMF, which systematically under-selected near $k = 2$ (Fig. \ref{fig:bo}E).




## Survival supervision and standard NMF produce different factor structures in PDAC {.unnumbered}

Bayesian optimization selected $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$ and $\alpha = `r round(tar_read(tar_params_best_tcgacptac)$alpha, 3)`$ for DeSurv in the PDAC training cohorts (TCGA and CPTAC). To compare the factor structures produced with and without survival supervision at the same dimensionality, we fit standard NMF at DeSurv's BO-selected rank ($k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$) and examined the overlap between each method's factor-specific gene rankings and established PDAC gene programs (Fig. \ref{fig:bio}A--B). Matching ranks ensures that differences in factor content reflect the presence or absence of survival supervision rather than differences in model complexity; whether NMF's performance improves at higher ranks is evaluated in the sensitivity analysis below and in SI Appendix, Table S4. We first describe the biological identity of each method's factors, then ask whether these structural differences correspond to differences in survival contribution.

We refer to the three DeSurv factors as D1--D3 and the three NMF factors as N1--N3. In the DeSurv factorization, D1 correlated with both classical tumor programs and an iCAF-associated stromal signature, D2 with activated immune-stromal programs, and D3 with basal-like tumor programs (Fig. \ref{fig:bio}A). The co-occurrence of classical and iCAF signatures within a single factor is consistent with evidence that survival stratification in PDAC improves when tumor-intrinsic and CAF subtypes are considered jointly [@peng2024determination; @elyada2019cross]. No DeSurv factor showed a significant positive correlation with exocrine-associated gene programs (DECODER Exocrine, Collisson Exocrine, Bailey ADEX; Fig. \ref{fig:bio}A). The survival contributions of these factors and their relationship to expression variance are quantified below (SI Appendix, Fig. S7).

Standard NMF produced a different factor organization (Fig. \ref{fig:bio}B). One factor (N2) was strongly and significantly correlated with exocrine-associated gene programs and negatively correlated with immune and stromal signatures (Fig. \ref{fig:bio}B). A second factor (N1) correlated with classical tumor identity (Fig. \ref{fig:bio}B), while a third (N3) correlated with immune, fibroblast, and extracellular matrix-related programs (including Puleo Desmoplastic, SCISSORS panCAF and myCAF, Maurer Immune and ECM, and DECODER Activated Stroma), aggregating these into a single factor rather than separating them as DeSurv does (D1 and D2; Fig. \ref{fig:bio}A). At $k = 3$, NMF thus devotes one of its three factors to exocrine-associated variation, leaving two factors to capture all tumor-intrinsic and microenvironmental programs. Whether these structural differences correspond to differences in prognostic information is quantified next.

We quantified per-factor survival contribution using two metrics (SI Appendix, Fig. S7; SI Methods). For each factor, we computed the conditional variance explained (semi-partial $R^2$: the incremental fraction of total variance accounted for by that factor given all other factors), and its survival contribution, measured as the change in Cox partial log-likelihood ($\Delta\ell$) when that factor is dropped from a $k$-factor Cox model. All three NMF factors had negligible $\Delta\ell$ despite spanning a range of variance, while DeSurv concentrated nearly all survival contribution into D1. Specifically, D1 explained $`r desurv_maxsurv_pct`$% of variance with $\Delta\ell = `r desurv_maxsurv_dl`$, while the highest-variance NMF factor explained $`r nmf_maxvar_pct`$% of expression variance but contributed only $\Delta\ell = `r nmf_maxvar_dl`$ to survival. The highest-variance DeSurv factor (D3, $`r desurv_maxvar_pct`$% of variance) also contributed minimally to survival after accounting for D1. Thus, in PDAC the highest-variance factors are not the most prognostic, consistent with tumor purity analyses across cancer types [@aran2015systematic].

To further characterize how the two factorizations relate, we examined pairwise W-matrix gene loading correlations (SI Appendix, Fig. S8). NMF's classical-associated factor (N1) showed the strongest correspondence with DeSurv's basal-like factor (D3), and NMF's microenvironmental factor (N3) mapped primarily to DeSurv's D2. NMF's exocrine-associated factor (N2) did not correspond strongly to any single DeSurv factor, with its strongest (though weak) correlation to D3, consistent with D3's high variance and low survival contribution. Together, the two methods share overlapping tumor and microenvironmental structure but differ in how exocrine-associated variation is represented: NMF dedicates a factor to it, whereas DeSurv does not. The key quantitative difference is that DeSurv concentrates survival contribution into a single factor ($\Delta\ell = `r desurv_maxsurv_dl`$), whereas no NMF factor achieves comparable survival contribution at the same rank. Whether this translates to improved generalization across independent cohorts is tested below.
```{r fig-pdac, fig.cap="Survival supervision produces different factor structures in PDAC. (A--B) Spearman correlations between factor gene rankings and established PDAC gene programs for DeSurv (A) and standard NMF (B) at $k = 3$. Asterisks denote significance after multiple testing correction; only correlations $> 0.2$ shown. (C) Pairwise correspondence between NMF and DeSurv factors, measured by Spearman rank correlation of W-matrix gene loadings across all genes. (D) Fraction of expression variance explained versus survival contribution ($\\Delta$ Cox partial log-likelihood upon factor removal) for each factor. DeSurv concentrates survival signal into D1 while the highest-variance NMF factor (N2, exocrine-associated) contributes negligibly to survival. \\label{fig:pdac}", fig.env='figure*', fig.pos='t', out.height= "7in", out.width='\\textwidth'}

library(cowplot)
library(ggplot2)

# Panels A-B: gene program heatmaps
tar_load(fig_gene_overlap_heatmap_desurv_tcgacptac)
tar_load(fig_gene_overlap_heatmap_std_desurvk_tcgacptac)

# Panel C: W-matrix correlation (promoted from supplement)
tar_load(fig_desurv_std_correlation_tcgacptac)

# Panel D: variance vs survival (promoted from supplement)
tar_load(fig_variation_explained_tcgacptac)

# ── Panels A-B: heatmaps with shared legend ──
plot_3a <- fig_gene_overlap_heatmap_desurv_tcgacptac$plot
plot_3b <- fig_gene_overlap_heatmap_std_desurvk_tcgacptac$plot
legend_ab <- fig_gene_overlap_heatmap_desurv_tcgacptac$legend
legend_ab <- gtable::gtable_add_padding(
  legend_ab, padding = unit(c(0, 10, 0, 0), "pt")
)

top_row <- plot_grid(
  plot_3a, plot_3b, cowplot::ggdraw(legend_ab),
  ncol = 3, labels = c("A", "B", ""), align = "hv",
  label_size = 12, rel_widths = c(2.5, 2.5, 0.6)
)

# ── Panel C: W-matrix correlation ──
plot_3c <- plot_grid(
  fig_desurv_std_correlation_tcgacptac$plot,
  plot_grid(
    NULL,
    cowplot::ggdraw(fig_desurv_std_correlation_tcgacptac$legend),
    nrow = 2, rel_heights = c(0.08, 0.92)
  ),
  ncol = 2, rel_widths = c(4, 1)
)

# ── Panel D: variance vs survival scatterplot ──
plot_3d <- set_fig_font(fig_variation_explained_tcgacptac, size = 10) +
  theme(
    legend.position    = c(1, 0.5),
    legend.justification = c(1, 0),
    legend.background  = element_rect(color = "black"),
    axis.title         = element_text(size = 8)
  )

bottom_row <- plot_grid(
  plot_3c, plot_3d,
  ncol = 2, labels = c("C", "D"),
  label_size = 12, rel_widths = c(1.2, 1)
)

plot_grid(top_row, bottom_row, nrow = 2, rel_heights = c(1.4, 1))
```


## Survival-aligned programs generalize across independent PDAC cohorts {.unnumbered}

A prognostic factorization is useful only if it generalizes beyond the training cohort. Having established that survival supervision and standard NMF produce different factor structures in the training data, we next tested whether DeSurv's structure transfers to independent cohorts without retraining. We computed factor scores in each validation dataset by projecting the gene programs learned in the training cohort ($Z = \tilde{W}^\top X_{\text{new}}$), yielding a continuous score for each factor per sample.

For DeSurv, we evaluated D1 (the classical--iCAF co-program with the largest training-data $\Delta\ell$); for standard NMF at the same rank, we selected the factor with the largest $\Delta\ell$ (N1, the classical-associated factor). External validation C-index for NMF at independently selected ranks is summarized in SI Appendix, Table S4; standard NMF at elbow-selected $k = 5$ and BO-selected $k = 7$ showed no systematic concordance advantage over NMF at $k = 3$, indicating that additional unsupervised factors do not recover the prognostic signal that supervision concentrates. Across `r n_val_cohorts` independent PDAC cohorts ($n = `r n_val_total`$), the DeSurv linear predictor showed a consistent survival association (stratified Cox model: pooled HR $`r hr_pool`$; 95% CI $`r ci_pool_lo`$--$`r ci_pool_hi`$; $P = `r p_pool`$; per-cohort forest plot in SI Appendix, Fig. S9; KM curves in Fig. S5C--F). The analogous NMF linear predictor showed a weaker pooled effect (HR $`r hr_pool_nmf`$; 95% CI $`r ci_pool_lo_nmf`$--$`r ci_pool_hi_nmf`$; $P = `r p_pool_nmf`$); a direct comparison using dichotomized risk groups is presented below.

To assess clinical interpretability of these continuous associations, we dichotomized validation samples into high- and low-risk groups using method-specific cutpoints, each independently optimized via cross-validation on the training data. The DeSurv linear predictor separated survival trajectories in the pooled validation cohort (HR $= `r km_desurv$hr`$, 95% CI $`r km_desurv$ci_lo`$--$`r km_desurv$ci_hi`$, $P `r if(km_desurv$p_lt) "< 0.001" else paste0("= ", km_desurv$p_val)`$; Fig. \ref{fig:bio}C). The same procedure applied to NMF yielded weaker stratification (HR $= `r km_nmf$hr`$, 95% CI $`r km_nmf$ci_lo`$--$`r km_nmf$ci_hi`$, $P `r if(km_nmf$p_lt) "< 0.001" else paste0("= ", km_nmf$p_val)`$; Fig. \ref{fig:bio}D). Together, these results indicate that incorporating survival information during factorization yields gene programs with more transportable prognostic associations than the standard discover-then-evaluate approach.


## Factorization rank K=3 is robust across supervision settings {.unnumbered}

```{r ksens-counts, echo=FALSE}
# ── Load K-sensitivity RDS outputs ────────────────────────────────────────────
# Generated by:
#   Rscript inst/cv_grid_validation_analysis_amber_optimal.R  → adj_p_270, hcor_270, master_rows_270
#   Rscript inst/cv_grid_validation_analysis_allgenes.R       → adj_p_all, hcor_all, master_rows_all
#   Rscript inst/k3_k7_nesting_heatmap.R                     → k3_k7_summary
#   Rscript inst/create_ksens_rds.R                          → lam300_summary, production_summary (bootstrap)
RDS_DIR <- "results/cv_grid"

adj_p_270    <- readRDS(file.path(RDS_DIR, "adj_p_270_matrix.rds"))
adj_p_all    <- readRDS(file.path(RDS_DIR, "adj_p_all_matrix.rds"))
unadj_p_270  <- readRDS(file.path(RDS_DIR, "unadj_p_270_matrix.rds"))
hcor_270     <- readRDS(file.path(RDS_DIR, "hcor_270_matrix.rds"))
hcor_all     <- readRDS(file.path(RDS_DIR, "hcor_all_matrix.rds"))
master_270   <- readRDS(file.path(RDS_DIR, "master_rows_270.rds"))
master_all   <- readRDS(file.path(RDS_DIR, "master_rows_all.rds"))
lam300       <- readRDS(file.path(RDS_DIR, "lam300_summary.rds"))
prod_sum     <- readRDS(file.path(RDS_DIR, "production_summary.rds"))

# ── Derived scalars from unadjusted p-value matrix ────────────────────────────
n_sig_270            <- sum(unadj_p_270 < 0.05)
n_k3_sig_270         <- sum(unadj_p_270["K=3", ] < 0.05)
n_k7_sig_270         <- sum(unadj_p_270["K=7", ] < 0.05)
k3_alpha55_p_270     <- unadj_p_270["K=3", "a=0.55"]
k7_alpha75_p_270     <- unadj_p_270["K=7", "a=0.75"]

# ── Derived scalars from adjusted p-value matrix ─────────────────────────────
n_k3_adj_sig_270     <- sum(adj_p_270["K=3", ] < 0.05)
n_k7_adj_sig_270     <- sum(adj_p_270["K=7", ] < 0.05)
n_sig_all            <- sum(adj_p_all < 0.05)
k3_alpha55_p_all     <- adj_p_all["K=3", "a=0.55"]
k5_alpha55_p_270     <- adj_p_270["K=5", "a=0.55"]
k5_alpha25_p_all     <- adj_p_all["K=5", "a=0.25"]

# ── H-cor values from hcor matrices ───────────────────────────────────────────
hcor_k3_a55_270  <- hcor_270["K=3", "a=0.55"]
hcor_k5_a55_270  <- hcor_270["K=5", "a=0.55"]
hcor_k5_a25_all  <- hcor_all["K=5", "a=0.25"]

# ── λ=0.300 analysis scalars ───────────────────────────────────────────────────
n_sig_all_prior  <- lam300$n_sig
production_adj_p <- prod_sum$adj_p

# ── Grid dimensions (for "of N combinations" and "of M alpha values" prose) ──
GRID_K      <- c(2, 3, 5, 7, 9)
GRID_ALPHA  <- c(0, 0.25, 0.35, 0.55, 0.75, 0.85, 0.95)
n_alpha     <- length(GRID_ALPHA)
n_grid      <- length(GRID_K) * length(GRID_ALPHA)
```

The 1-SE model selection rule chose $k = `r tar_read(tar_k_selection_tcgacptac)$k_selected`$ over the global BO maximum ($k = `r tar_read(tar_k_selection_tcgacptac)$k_best`$; CV C-index $`r k_best_cindex`$ versus $`r k3_best_cindex`$; margin $`r k_margin`$). To assess whether this reflects a robust solution or an artifact of a specific tuning configuration, we conducted a sensitivity analysis across $K \in \{2, 3, 5, 7, 9\}$ and $\alpha \in \{0, 0.25, 0.35, 0.55, 0.75, 0.85, 0.95\}$ (`r n_grid` combinations), holding all other hyperparameters at their BO-selected values and evaluating each combination in independent external validation cohorts via stratified Cox regression (SI Appendix, Tables S1--S3). $K = 3$ achieved significance ($P < 0.05$) across `r n_k3_sig_270` of `r n_alpha` supervision strengths, and $K = `r tar_read(tar_k_selection_tcgacptac)$k_best`$ at a comparable `r n_k7_sig_270` of `r n_alpha`. After additional adjustment for PurIST and DeCAF molecular classifiers, $K = 3$ retained significance at `r n_k3_adj_sig_270` of `r n_alpha` supervision strengths, while $K = `r tar_read(tar_k_selection_tcgacptac)$k_best`$ dropped to `r n_k7_adj_sig_270` of `r n_alpha` (SI Appendix, Table S2), indicating that much of $K = `r tar_read(tar_k_selection_tcgacptac)$k_best`$'s prognostic signal overlaps with existing molecular classifiers. The parsimonious $K = 3$ solution thus validates robustly across supervision strengths and retains prognostic value beyond known classifiers, providing empirical support for the 1-SE rule.
