---
output:
  pdf_document: default
  html_document: default
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)
library(targets)
if (!exists("set_fig_font", inherits = TRUE)) {
  set_fig_font <- function(plot_obj, size = 10) {
    if (inherits(plot_obj, "ggplot")) {
      plot_obj + ggplot2::theme(text = ggplot2::element_text(size = size))
    } else {
      plot_obj
    }
  }
}
```

## The NMF–Cox model shows consistent convergence and stability across restarts

Across datasets and initialization schemes, the NMF–Cox algorithm consistently converged to numerically stable solutions within the designated iteration budget. Warm-start strategies, in which solutions at $\alpha=0$ initialized supervised runs, substantially reduced variability across restarts and improved reproducibility. Figure \@ref(fig:fig-converge) shows representative loss trajectories demonstrating monotone decreases until convergence across restarts. Compared with naïve random initialization, warm-starts produced tighter distributions of cross-validated C-index and partial likelihood, confirming stability of the optimization procedure.

```{r fig-converge, fig.width = 6, fig.height= 4, fig.cap="Convergence of model across initializations." , fig.env='figure*', fig.pos='t', out.height= "4in", out.width='\\textwidth'}
tar_load(desurv_seed_fits_tcgacptac)
set.seed(147)
seeds = sample(1:length(desurv_seed_fits_tcgacptac$fits),5,replace = FALSE)
iters <- list()
for (i in seeds) {
  loss_vec <- desurv_seed_fits_tcgacptac$fits[[i]]$lossit
  iters[[i]] <- data.frame(
    lossit = loss_vec,
    slossit = desurv_seed_fits_tcgacptac$fits[[i]]$slossit,
    nlossit = desurv_seed_fits_tcgacptac$fits[[i]]$nlossit,
    iter = seq_along(loss_vec),
    init = i
  )
}

lossit <- dplyr::bind_rows(iters) |> dplyr::filter(iter < 5000)

lossit = lossit %>%
  group_by(init) %>%
  arrange(iter) %>%
  mutate(
    rel_delta_loss = lossit / dplyr::first(lossit),
    rel_delta_sloss = slossit / dplyr::first(slossit),
    rel_delta_nloss = nlossit / dplyr::first(nlossit)
  ) %>%
  filter(!is.na(rel_delta_loss))

ggplot2::ggplot(lossit, ggplot2::aes(y = rel_delta_loss, x = iter, color = as.factor(init))) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::labs(x = "Iteration", y = "Loss normalized to initial value", color = "Initialization")

```


## Simulation results under null and mixed scenarios {.unnumbered}

To characterize the operating characteristics of DeSurv across different signal regimes, we evaluated performance under three simulation scenarios. In the primary scenario (R0_easy; main text Fig. 3), prognostic programs explained low variance relative to outcome-neutral background signals, and survival depended on 150 factor-specific marker genes ($\beta_1 = 2$, $\beta_2 = \beta_3 = 0$). In the null scenario (R00_null), survival times were generated independently of gene expression ($\beta = 0$), so no prognostic structure existed. In the mixed scenario (R_mixed), survival depended on 300 genes per lethal factor, of which 50% were factor-specific markers and 50% were background genes with loadings shared across all factors. All scenarios used $G = 3{,}000$ genes, $N = 200$ samples, $K = 3$ true factors, and identical gamma distribution parameters for the gene loading matrix $W$ (see Materials and Methods). Performance was evaluated using cross-validated concordance index (C-index), precision of recovered prognostic gene sets, and the distribution of selected ranks across simulation replicates.

Under the null scenario (Fig. S\ref{fig:sim-null-mixed}A--B), DeSurv showed no advantage over unsupervised NMF ($\alpha = 0$). Both methods yielded C-index values near 0.5, consistent with the absence of survival signal. The distribution of selected ranks showed similar variability across methods. These results confirm that DeSurv does not hallucinate prognostic structure when none exists: Bayesian optimization selected low values of the supervision parameter $\alpha$, effectively recovering standard NMF as a special case. This specificity control is important because it establishes that the advantages observed in the primary scenario reflect genuine signal recovery rather than overfitting to noise in the survival labels.

Under the mixed scenario (Fig. S\ref{fig:sim-null-mixed}C--E), DeSurv's advantage over unsupervised NMF was present but attenuated relative to the primary scenario. C-index values were modestly higher for DeSurv, and precision for recovering the true survival gene set---which now included both marker and background genes---showed a smaller improvement compared to the primary scenario. This attenuation is expected: when survival-relevant genes partially overlap with variance-dominant background programs, unsupervised NMF captures some of the prognostic structure incidentally, reducing the marginal benefit of supervision. The mixed scenario thus identifies the intermediate regime where DeSurv offers a moderate but not dramatic improvement over unsupervised approaches.

Together, the three scenarios provide a dose-response relationship: DeSurv's advantage is greatest when variance and prognosis diverge (primary), moderate when they partially overlap (mixed), and absent when no survival signal exists (null). This gradient is consistent with the predictions of sufficient dimension reduction theory and offers practical guidance for users: DeSurv is most beneficial in settings where the dominant transcriptional axes are expected to be prognostically neutral, as is common in cancers with high tumor purity variation or dominant tissue-composition signals.

```{r fig-sim-null-mixed, fig.width = 6.5, fig.height = 6, fig.cap = "DeSurv performance under null and mixed simulation scenarios. (A--B) Null scenario: survival times are independent of gene expression ($\\beta = 0$). DeSurv shows no advantage over unsupervised NMF ($\\alpha = 0$), confirming that the method does not hallucinate prognostic structure when none exists. (A) C-index distributions are comparable and centered near 0.5. (B) Distribution of selected ranks shows comparable variability; both methods favor parsimonious solutions. Precision is not shown because no true survival genes exist under the null ($\\beta = 0$). (C--E) Mixed scenario: survival depends on a mixture of factor-specific marker genes (50\\%) and shared background genes (50\\%), creating partial overlap between variance-dominant and prognostically relevant structure. DeSurv's advantage is present but attenuated relative to the primary scenario (main text Fig. 3). (C) C-index is modestly higher for DeSurv. (D) Precision for recovering the true survival gene set (now including both marker and background genes) remains substantially higher for DeSurv. (E) Selected-rank distributions. In all panels, DeSurv (blue) denotes the supervised model and NMF (red) denotes the unsupervised baseline ($\\alpha = 0$), both tuned via Bayesian optimization. \\label{fig:sim-null-mixed}", fig.env='figure*', fig.pos='t', out.height = "6in", out.width='\\textwidth'}
library(targets)
library(cowplot)
library(ggplot2)

tar_load(sim_figs_by_scenario)

scenario_ids <- sapply(sim_figs_by_scenario, function(x) x$scenario_id)
analysis_ids <- sapply(sim_figs_by_scenario, function(x) x$analysis_id)

null_idx <- which(scenario_ids == "R00_null" & analysis_ids == "bo_tune_ntop")
mixed_idx <- which(scenario_ids == "R_mixed" & analysis_ids == "bo_tune_ntop")

null_plots <- sim_figs_by_scenario[[null_idx]]
mixed_plots <- sim_figs_by_scenario[[mixed_idx]]

null_plots <- lapply(null_plots, set_fig_font, size = 10)
mixed_plots <- lapply(mixed_plots, set_fig_font, size = 10)

# Null row: C-index and k-hist only (precision undefined when beta = 0)
null_cindex <- null_plots$cindex_box +
  labs(title = NULL) +
  scale_y_continuous(limits = c(0.3, 1))
null_cindex$data$scenario_id <- "Null scenario (\u03b2 = 0)"

null_khist <- null_plots$k_hist + labs(title = NULL)
null_khist$data$scenario_id <- "Null scenario (\u03b2 = 0)"

null_row <- plot_grid(
  null_cindex, null_khist,
  ncol = 2, labels = c("A", "B"), rel_widths = c(1, 1.2)
)

# Mixed row: C-index, precision, and k-hist
mixed_cindex <- mixed_plots$cindex_box +
  labs(title = NULL) +
  scale_y_continuous(limits = c(0.3, 1))
mixed_cindex$data$scenario_id <- "Mixed scenario"

mixed_prec <- mixed_plots$precision_box + labs(title = NULL)
mixed_prec$data$scenario_id <- "Mixed scenario"

mixed_khist <- mixed_plots$k_hist + labs(title = NULL)
mixed_khist$data$scenario_id <- "Mixed scenario"

mixed_row <- plot_grid(
  mixed_cindex, mixed_prec, mixed_khist,
  ncol = 3, labels = c("C", "D", "E")
)

plot_grid(null_row, mixed_row, nrow = 2, rel_heights = c(1, 1))
```


## Cross-validated C-index across factorization rank {.unnumbered}

To evaluate how the number of latent factors $k$ affects prognostic performance, we performed an exhaustive grid search over $k \in \{2, 3, \ldots, 12\}$ and supervision strength $\alpha \in \{0, 0.05, \ldots, 0.95\}$ using 5-fold cross-validation on the TCGA+CPTAC training cohort. For each $k$, the best-performing $\alpha$ was selected by maximizing the mean CV C-index. The unsupervised NMF baseline ($\alpha = 0$) was evaluated at the same $k$ values for comparison. Results are shown for two gene-selection settings: ntop = 270 top genes per factor, and ntop = ALL (all genes retained), both at $\lambda = 0.349$ and $\nu = 0.056$.

Figure \@ref(fig:fig-cindex-by-k) shows the resulting C-index curves for both training (CV) and external validation (pooled across held-out cohorts), with shaded ribbons indicating $\pm 1$ standard error. In training, DeSurv consistently outperforms NMF across all $k$, confirming that survival supervision improves in-sample concordance. In external validation, DeSurv maintains a comparable or improved C-index relative to NMF, demonstrating that the supervised signal generalizes beyond the training data rather than reflecting overfitting to the survival labels.

```{r fig-cindex-by-k, fig.width = 10, fig.height = 10, fig.cap="Training (CV) and external validation C-index as a function of factorization rank $k$ for DeSurv (blue) and standard NMF (red, $\\alpha = 0$). Shaded ribbons show $\\pm 1$ SE. For each $k$, the best $\\alpha$ was selected by maximizing the mean CV C-index. Validation C-index is pooled across all external cohorts. (A) Results using ntop = 270 top genes per factor ($\\lambda = 0.349$, $\\nu = 0.056$). (B) Results retaining all genes (ntop = ALL; $\\lambda = 0.349$, $\\nu = 0.056$).", fig.env='figure*', fig.pos='t', out.width='\\textwidth', echo=FALSE}
library(magick)
library(cowplot)
# This figure is produced by the cv_grid pipeline (separate store);
# read directly from the file output.
fig_path <- "figures/cv_grid/cv_cindex_by_k_primary.pdf"
img <- magick::image_read_pdf(fig_path, density = 300)
cowplot::ggdraw() + cowplot::draw_image(img)
```

## K-sensitivity analysis: effect of gene focusing on iCAF factor validation {.unnumbered}

To evaluate the robustness of the K=3 iCAF factorization, we conducted an exhaustive sensitivity analysis across $K \in \{2, 3, 5, 7, 9\}$ and $\alpha \in \{0, 0.25, 0.35, 0.55, 0.75, 0.85, 0.95\}$ (35 combinations), holding all remaining hyperparameters fixed at the production-optimal values ($\lambda = 0.349$, $\nu = 0.056$, 100 initializations). For each fit, we identified the factor most correlated with the production K=3 Factor 1 (the iCAF factor) and evaluated its prognostic value in 570 independent validation patients across four cohorts (Dijk, Moffitt GEO, Puleo, PACA-AU), using stratified Cox regression (\texttt{strata(dataset)}) with and without adjustment for PurIST and DeCAF classifiers. Factor-program fidelity was quantified by Spearman rank correlation of ntop=270-focused H-scores (H-cor) with the production factor. We repeated this analysis under two gene-projection settings --- ntop=270 (production-optimal top genes per factor) and ntop=NULL (all 1,970 genes) --- to isolate the effect of gene focusing on validation performance.

The central finding is shown in Table \ref{tab:tab-ksens-summary}: **ntop=270 is the critical hyperparameter enabling the production fit's adjusted validation.** Moving from ntop=NULL to ntop=270, while holding $\lambda$ and $\nu$ at their optimal values, increases the number of K $\times$ $\alpha$ combinations achieving adjusted $p < 0.05$ from 5/35 to 16/35 and restores K=3 at $\alpha=0.55$ to adjusted $p = 0.003$ --- nearly identical to the production fit ($p = 0.004$). By contrast, changing only $\lambda$ from 0.3 to 0.349 while keeping ntop=NULL leaves the overall count unchanged (5/35), confirming that $\lambda/\nu$ tuning alone cannot reproduce the adjusted signal.

```{r tab-ksens-summary, echo=FALSE, results='asis'}
library(knitr)
library(kableExtra)

tab_summary <- data.frame(
  Condition = c(
    "ntop = NULL, $\\lambda$ = 0.300, $\\nu$ = 0.050",
    "ntop = NULL, $\\lambda$ = 0.349, $\\nu$ = 0.056",
    "ntop = 270,\\hspace{2pt} $\\lambda$ = 0.349, $\\nu$ = 0.056"
  ),
  Count = c("5 / 35", "5 / 35", "\\textbf{16 / 35}"),
  K3_p  = c("0.072", "0.097", "\\textbf{0.003}"),
  Best  = c(
    "K=7, $\\alpha$=0.75 (adj $p$ = 0.019)",
    "K=5, $\\alpha$=0.25 (adj $p$ = 0.0003)",
    "K=3, $\\alpha$=0.55 (adj $p$ = 0.003)"
  ),
  stringsAsFactors = FALSE
)
colnames(tab_summary) <- c(
  "Analysis condition",
  "Adj $p < 0.05$ (of 35)",
  "K=3, $\\alpha$=0.55 adj $p$",
  "Best single combination"
)

kable(tab_summary, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = "Three-way comparison of K-sensitivity analyses differing only in gene-projection setting (ntop) and regularization ($\\lambda$, $\\nu$). All analyses span K $\\in$ \\{2, 3, 5, 7, 9\\} $\\times$ $\\alpha \\in$ \\{0, 0.25, 0.35, 0.55, 0.75, 0.85, 0.95\\} = 35 combinations. Count shows combinations achieving adjusted $p < 0.05$; bold highlights the ntop=270 result that restores K=3 adjusted significance.",
      label = "tab-ksens-summary") |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

Tables \ref{tab:tab-ksens-adjp-ntop270} and \ref{tab:tab-ksens-adjp-allgenes} present the full K $\times$ $\alpha$ adjusted p-value matrices for the ntop=270 and ntop=NULL analyses respectively. With ntop=270, K=3 validates adjustedly at 5 of 7 $\alpha$ values tested ($\alpha \in \{0.00, 0.25, 0.55, 0.75, 0.95\}$), and broad adjusted significance is also observed at K=2 (5/7) and K=5 (4/7). With ntop=NULL (Table \ref{tab:tab-ksens-adjp-allgenes}), the landscape is sparse and structurally different: K=3 at $\alpha=0.55$ narrowly misses significance ($p = 0.097$), and only isolated combinations scattered across K=3, K=5, K=7, and K=9 achieve adjusted $p < 0.05$.

```{r tab-ksens-adjp-ntop270, echo=FALSE, results='asis'}
fmt_p <- function(x) {
  s <- ifelse(x < 0.001,
    formatC(x, format = "e", digits = 0),
    formatC(x, format = "f", digits = 3)
  )
  s <- sub("e-0(\\d)$", "e-\\1", s)
  ifelse(x < 0.05, paste0("\\textbf{", s, "}"), s)
}

alpha_labs <- c("0.00", "0.25", "0.35", "0.55", "0.75", "0.85", "0.95")
k_labs     <- c("K=2", "K=3", "K=5", "K=7", "K=9")

adj_p_270 <- matrix(c(
  0.564, 0.038, 0.038, 0.019, 0.060, 0.011, 0.002,
  0.030, 0.030, 0.519, 0.003, 0.028, 0.263, 0.009,
  5e-4,  0.005, 8e-4,  0.019, 0.073, 0.604, 0.447,
  0.513, 0.471, 0.001, 0.041, 0.676, 0.131, 0.130,
  3e-4,  0.299, 0.243, 0.005, 0.943, 0.867, 0.263
), nrow = 5, byrow = TRUE)

df_270 <- as.data.frame(
  matrix(fmt_p(adj_p_270), nrow = 5, dimnames = list(k_labs, alpha_labs))
)

kable(df_270, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = "Adjusted validation p-values (Cox + PurIST + DeCAF + strata(dataset)) across K $\\times$ $\\alpha$ with ntop=270, $\\lambda=0.349$, $\\nu=0.056$. Each cell gives the p-value for the factor most correlated with the production K=3 iCAF factor. Bold: $p < 0.05$; 16 of 35 combinations are significant.",
      label = "tab-ksens-adjp-ntop270") |>
  kable_styling(latex_options = c("hold_position"), font_size = 9) |>
  add_header_above(c(" " = 1, "$\\\\alpha$" = 7), escape = FALSE)
```

```{r tab-ksens-adjp-allgenes, echo=FALSE, results='asis'}
adj_p_all <- matrix(c(
  0.414, 0.192, 0.142, 0.197, 0.090, 0.194, 0.158,
  0.205, 0.024, 0.280, 0.097, 0.112, 0.076, 0.111,
  0.461, 3e-4,  0.150, 0.233, 0.067, 0.124, 0.197,
  0.394, 0.336, 0.304, 0.097, 0.094, 0.062, 0.021,
  0.023, 0.093, 0.474, 0.022, 0.106, 0.127, 0.122
), nrow = 5, byrow = TRUE)

df_all <- as.data.frame(
  matrix(fmt_p(adj_p_all), nrow = 5, dimnames = list(k_labs, alpha_labs))
)

kable(df_all, format = "latex", escape = FALSE, booktabs = TRUE,
      caption = "Adjusted validation p-values across K $\\times$ $\\alpha$ with ntop=NULL (all 1,970 genes), $\\lambda=0.349$, $\\nu=0.056$. Layout identical to Table \\ref{tab:tab-ksens-adjp-ntop270}. Only 5 of 35 combinations are significant (bold), and K=3 at $\\alpha=0.55$ (the production neighborhood) reaches only $p = 0.097$.",
      label = "tab-ksens-adjp-allgenes") |>
  kable_styling(latex_options = c("hold_position"), font_size = 9) |>
  add_header_above(c(" " = 1, "$\\\\alpha$" = 7), escape = FALSE)
```

Four additional findings emerge from these analyses. First, K=7's apparent advantage observed in a parallel analysis using conventional hyperparameters ($\lambda=0.3$, ntop=NULL), where K=7 at $\alpha=0.75$ achieved adjusted $p = 0.019$, completely disappears with ntop=270 (adj $p = 0.676$, Table \ref{tab:tab-ksens-adjp-ntop270}). Focused 270-gene projection fragments the iCAF signal across K=7's factors, establishing that the earlier K=7 advantage was an artifact of the all-genes projection rather than an intrinsic property of the 7-factor solution. Second, K=3 robustly validates adjustedly across a broad $\alpha$ range with ntop=270 (5/7 values), demonstrating that the production result does not depend on the BO-selected $\alpha = 0.334$ specifically. Third, K=5 at $\alpha=0.55$ achieves the highest factor-program fidelity in the ntop=270 analysis (H-cor = 0.903 vs. K=3's 0.855) and also validates adjustedly ($p = 0.019$), providing convergent evidence for the iCAF biology from a higher-K model; an independent K=5 all-genes fit at $\alpha=0.25$ likewise validates adjustedly ($p = 0.0003$, H-cor = 0.910). Fourth, without gene focusing, high supervision weights ($\alpha \geq 0.75$) produce factors with extreme training discrimination (HR $\approx 0.07$–$0.10$, LRT $p \sim 10^{-60}$) that fail external validation, indicating overfitting of the training survival signal; gene focusing (ntop=270) suppresses this by constraining each factor to its 270 most biologically characteristic genes.

## K=3 factors are partially but asymmetrically nested within K=7 {.unnumbered}

To assess whether the K=3 solution is simply a pruned version of the K=7 solution, we computed pairwise Spearman correlations between the W-matrix gene loadings of the production K=3 model and a K=7 DeSurv fit at $\alpha = 0.35$ (the nearest grid value to the BO-selected optimum at $\alpha = 0.33$), both using ntop$=270$, $\lambda = 0.349$, $\nu = 0.056$.

The correspondence was partial but asymmetric (Fig. \ref{fig:k3-k7-nesting}). Two of three K=3 factors had clear one-to-one K=7 counterparts: K3\_F1 mapped strongly to K7\_F1 ($r = 0.77$) and K3\_F2 mapped very strongly to K7\_F7 ($r = 0.88$). In contrast, K3\_F3 showed no single dominant K=7 match; instead, it correlated moderately with four K=7 factors simultaneously (K7\_F2 through K7\_F6; $r = 0.57$--$0.71$). K7\_F4 showed weak-to-moderate correlations with all three K=3 factors ($r = 0.45$--$0.52$), suggesting it is a mixed composite with no clear biological counterpart in the K=3 solution.

This fragmentation pattern indicates that K=7 does not simply subdivide K=3 structure. Rather, it disperses one coherent K=3 prognostic program across multiple lower-signal components, consistent with the observed failure of K=7 to validate adjustedly in external cohorts (adjusted $p = 0.676$; see main text). The K=3 solution concentrates biological signal that K=7 dilutes.

```{r fig-k3-k7-nesting, fig.width = 7, fig.height = 3.5, fig.cap = "Factor correspondence between the production K=3 model and a K=7 DeSurv fit ($\\alpha = 0.35$, ntop$=270$), measured by Spearman rank correlation of W-matrix gene loadings across all genes. Two K=3 factors have clear K=7 counterparts (K3\\_F1 $\\to$ K7\\_F1, $r = 0.77$; K3\\_F2 $\\to$ K7\\_F7, $r = 0.88$), while K3\\_F3 is distributed broadly across four K=7 factors ($r = 0.57$--$0.71$), indicating fragmentation rather than nesting. \\label{fig:k3-k7-nesting}", fig.env='figure*', fig.pos='t', out.width='\\textwidth', echo=FALSE}
library(magick)
library(cowplot)
img <- magick::image_read_pdf("figures/cv_grid/k3_k7_factor_correlation_heatmap.pdf", density = 300)
cowplot::ggdraw() + cowplot::draw_image(img)
```

## Cutpoint selection and validation of dichotomized risk groups {.unnumbered}

To convert the continuous DeSurv linear predictor into a clinically interpretable binary risk stratification, we selected an optimal z-score cutpoint via cross-validation on the TCGA+CPTAC training cohort. For each candidate cutpoint in $\{-2.0, -1.8, \ldots, 2.0\}$, we computed the mean absolute log-rank z-statistic across held-out folds and selected the cutpoint maximizing this metric (Figure \@ref(fig:fig-cutpoint-km)A). We then applied this cutpoint to external validation cohorts by standardizing each patient's linear predictor using the training mean and standard deviation. Figure \@ref(fig:fig-cutpoint-km)B shows the resulting Kaplan–Meier curves for the pooled external validation analysis, with a stratified log-rank test accounting for dataset of origin. Panels C–F display per-cohort validation KM curves for Dijk, Moffitt, PACA-AU, and Puleo, respectively. Across all cohorts, the DeSurv-derived high-risk group showed consistently shorter survival, confirming that the training-derived cutpoint generalizes to independent datasets.

```{r fig-cutpoint-km, fig.width = 10, fig.height = 14, fig.cap = "Cutpoint selection and external validation of dichotomized risk groups. (A) Cross-validated mean absolute log-rank z-statistic as a function of z-score cutpoint; red dashed line indicates the selected optimum. (B) Pooled validation Kaplan--Meier curves (stratified log-rank test). (C--D) Per-cohort validation KM curves: (C) Dijk, (D) Moffitt. (E--F) Per-cohort validation KM curves: (E) PACA-AU, (F) Puleo.", fig.env='figure*', fig.pos='t', out.width='\\textwidth', echo=FALSE}
library(cowplot)
library(magick)

# Load file targets to establish dependency chain
tar_load(fig_cutpoint_curves_tcgacptac)
tar_load(fig_val_km_dichot_tcgacptac)

read_pdf_as_grob <- function(path) {
  img <- magick::image_read_pdf(path, density = 300)
  cowplot::ggdraw() + cowplot::draw_image(img)
}

# fig_cutpoint_curves_tcgacptac is a character vector of file paths
cutpoint_logrank_path <- grep("logrank", fig_cutpoint_curves_tcgacptac, value = TRUE)
p_cutpoint <- read_pdf_as_grob(cutpoint_logrank_path)

# fig_val_km_dichot_tcgacptac is a character vector of file paths
km_paths <- fig_val_km_dichot_tcgacptac
p_pooled  <- read_pdf_as_grob(grep("km_val_pooled_logrank", km_paths, value = TRUE))
p_dijk    <- read_pdf_as_grob(grep("km_val_Dijk_logrank", km_paths, value = TRUE))
p_moffitt <- read_pdf_as_grob(grep("km_val_Moffitt_GEO_array_logrank", km_paths, value = TRUE))
p_paca    <- read_pdf_as_grob(grep("km_val_PACA_AU_logrank", km_paths, value = TRUE))
p_puleo   <- read_pdf_as_grob(grep("km_val_Puleo_array_logrank", km_paths, value = TRUE))

row1 <- plot_grid(
  p_cutpoint, p_pooled,
  ncol = 2, labels = c("A", "B"), label_size = 14
)

row2 <- plot_grid(
  p_dijk, p_moffitt,
  ncol = 2, labels = c("C", "D"), label_size = 14
)

row3 <- plot_grid(
  p_paca, p_puleo,
  ncol = 2, labels = c("E", "F"), label_size = 14
)

plot_grid(row1, row2, row3, nrow = 3, rel_heights = c(1, 1, 1))
```

## Overlap of DeSurv risk groups with known molecular subtypes {.unnumbered}

To assess whether the DeSurv-derived risk stratification captures biologically meaningful transcriptional programs, we examined the overlap between dichotomized risk groups (High vs Low) and two established PDAC molecular classifiers: PurIST (Basal-like vs Classical) and DeCAF (proCAF vs restCAF). Figure \@ref(fig:fig-subtype-overlap) shows the composition of each risk group across all pooled external validation cohorts, with Fisher's exact test p-values quantifying the association. The high-risk group was significantly enriched for Basal-like (PurIST) and proCAF (DeCAF) subtypes, consistent with the known poor prognosis of these molecular classes. This concordance confirms that DeSurv's survival-driven factorization recovers clinically relevant biology without requiring subtype labels during training.

```{r fig-subtype-overlap, fig.width = 8, fig.height = 5, fig.cap = "Subtype composition of DeSurv risk groups in pooled external validation cohorts. Stacked bar charts show the proportion of PurIST (left) and DeCAF (right) subtypes within Low and High risk groups defined by the log-rank-optimized cutpoint. Numbers indicate sample counts; Fisher's exact test p-values are shown above each panel.", fig.env='figure*', fig.pos='t', out.width='\\textwidth'}
library(magick)
library(cowplot)

tar_load(fig_subtype_overlap_tcgacptac)

overlap_path <- grep("subtype_overlap_pooled_logrank", fig_subtype_overlap_tcgacptac, value = TRUE)
img <- magick::image_read_pdf(overlap_path, density = 300)
cowplot::ggdraw() + cowplot::draw_image(img)
```

## Analysis of scRNA-seq {.unnumbered}
Next, we verify our findings at the cellular level using the Elyada scRNA-seq data [@elyada2019cross]. We found that our DeSurv factor 1 signature was expressed primarily in iCAF and B cells, factor 2 was expressed in Acinar and Classical PDAC 2, and factor 3 was expressed primarily in Basal-like PDAC cells (Figure \@ref(fig:fig-sc)B-E). This is consistent with our ORA analysis which found that factor 1 was mostly immune and factor 3 was basal-like, and with the overlap we saw with published gene lists that identified factor 1 as iCAF/classical/immune and factor 3 as basal-like. Interestingly, we do not see a large overlap of factor 1 with the classical PDAC cell types despite some overlap with classical gene lists. Need more here...

```{r fig-sc, fig.cap="A. Cell type clusters from the Elyada scRNA-seq data. B-D. VAM scores for the DeSurv factor signatures in each cell shown on the UMAP of cells in the Elyada-sc data. E. Heatmap of average VAM scores by cell type and factor signature in the Elyada-sc data.", fig.env='figure*', fig.pos='t', out.height= "8in", out.width='\\textwidth'}

tar_load(fig_sc_plot_tcgacptac)
fig_sc_plot_tcgacptac
```

## Variance explained versus survival contribution per factor {.unnumbered}

For each factor learned by DeSurv and standard NMF at $k = 3$, we compared the fraction of transcriptional variance explained with the Type III partial likelihood contribution to survival. The NMF factor explaining the greatest proportion of expression variance contributed minimally to survival prediction, consistent with its enrichment for exocrine and composition-driven programs. In contrast, the DeSurv factor with the largest survival contribution explained substantially less expression variance, directly illustrating the misalignment between variance-dominant and prognosis-relevant signals in PDAC (Fig. \ref{fig:varsurvival}).

```{r fig-varsurvival, fig.width = 4, fig.height = 3.5, fig.cap = "Fraction of expression variance explained versus Type III partial likelihood contribution for each factor ($k = 3$). Each point represents one factor; DeSurv factors (blue) and standard NMF factors (red) are labeled D1--D3 and N1--N3 respectively. Survival contribution is quantified as the reduction in Cox partial log-likelihood when each factor is removed from the full $k$-factor model (see SI Methods). \\label{fig:varsurvival}", fig.env='figure', fig.pos='t'}
tar_load(fig_variation_explained_tcgacptac)
fig_variation_explained_tcgacptac <- set_fig_font(fig_variation_explained_tcgacptac, size = 10)
fig_variation_explained_tcgacptac +
  theme(
    legend.position    = c(1, 0.5),
    legend.justification = c(1, 0),
    legend.background  = element_rect(color = "black"),
    axis.title         = element_text(size = 8)
  )
```

## DeSurv and NMF factor correspondence {.unnumbered}

Pairwise Spearman correlations between DeSurv and standard NMF factor scores illustrate how the two factorizations relate. Tumor-intrinsic structure is largely preserved across methods, with one DeSurv factor showing strong correspondence to the classical tumor-associated NMF factor. The microenvironmental NMF factor maps primarily to a single DeSurv factor enriched for activated immune and stromal programs. The exocrine-dominated NMF factor does not correspond strongly to any single DeSurv factor, consistent with the suppression of differentiation-driven signals under survival-informed optimization (Fig. \ref{fig:corr}).

```{r fig-corr, fig.width = 5, fig.height = 3.5, fig.cap = "Pairwise correspondence between NMF and DeSurv factors, measured by Spearman rank correlation of factor scores across training samples. Tumor-intrinsic structure is preserved across methods, while variance-driven microenvironmental structure is reorganized into a survival-aligned axis; the exocrine-dominated NMF factor shows no strong one-to-one correspondence with any DeSurv factor. \\label{fig:corr}", fig.env='figure', fig.pos='t'}
library(cowplot)
tar_load(fig_desurv_std_correlation_tcgacptac)
cowplot::plot_grid(
  fig_desurv_std_correlation_tcgacptac$plot,
  cowplot::ggdraw(fig_desurv_std_correlation_tcgacptac$legend),
  ncol = 2, rel_widths = c(4, 1)
)
```

## Per-cohort hazard ratios for the most prognostic factor {.unnumbered}

Hazard ratios for the factor with the largest training-data Cox partial log-likelihood improvement were estimated separately in each external validation cohort. DeSurv-derived factor hazard ratio estimates are more stable across cohorts than those from the corresponding standard NMF factor (Fig. \ref{fig:forest}).

```{r fig-forest, fig.width = 5, fig.height = 3.5, fig.cap = "Forest plot summarizing per-cohort hazard ratios (HRs; 95\\% CIs) for the latent factor most predictive of survival in the training data, evaluated in held-out validation datasets. Estimates from DeSurv (blue) show more consistent and predominantly protective associations across cohorts compared with standard NMF (red). \\label{fig:forest}", fig.env='figure', fig.pos='t'}
library(cowplot)
library(ggplot2)
tar_load(fig_hr_forest_tcgacptac)
base_size <- 10
theme_pnas <- theme_classic(base_size = base_size) +
  theme(
    plot.margin = margin(6, 10, 6, 6),
    legend.box.margin = margin(0, 0, 0, 0)
  )
fig_hr_forest_tcgacptac +
  scale_y_discrete(
    labels = c(
      "Puleo_array"       = "Puleo",
      "Dijk"              = "Dijk",
      "PACA_AU_seq"       = "PACA seq",
      "PACA_AU_array"     = "PACA array",
      "Moffitt_GEO_array" = "Moffitt"
    )
  ) +
  theme_pnas +
  theme(
    legend.position = "bottom",
    axis.title.y = element_blank()
  )
```
